---
title: 'Stats 250: Lab 3 Resources'
author: "Instructional Team"
date: "Monday, May 15"
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Learning Objectives
1. Explore how statistics change from sample to sample
2. Understand sampling distributions
3. Simulate the sampling distribution of Xbar for various sample sizes
4. Understand and apply the Central Limit Theorem (CLT)

### Functions 
1. `hist()`
2. `randomSample()`
3. `samplingDistXbar()`


***


## R Packages

We have used a lot of helpful functions in the first few labs. So far, all of these functions have been built into R. One powerful benefit of R is that users can create their own functions, store them in a package, and make it available for others to use. 

We have created our own package (called "stats250package") which contains useful functions that we will be utilizing over the next few labs. We will also be utilizing two other external packages. In order to load these packages, we need to run the following code chunk:

```{r loadPackages}
library(ggplot2)
library(ggdist)
library(stats250package)
```

Great - now we are ready to start the lab!


## Statistical Review

In lecture, you have learned about the following terms:

- Parameter: a summary value calculated from a population
- Statistic: a summary value calculated from a sample

The parameter is a fixed quantity that summarizes the entire population of interest. But what about the statistic - specifically the sample mean? You have discussed some of the following ideas in lecture:

- The sample mean is the average value of a random sample of observations
- The sample mean will vary from sample to sample
- This variability will decrease if the size of the sample is increased
- The sampling distribution of the sample mean helps us visualize how we expect a sample mean to vary from sample to sample

And this is what we will be illustrating in today's lab!


***


## Coding Examples

### The Population

This week we will be using the "employee" data set, which contains information for ***all*** 473 employees at a certain company. The variables included in this data set are:

- `salary`: annual salary (in dollars)
- `age`: age of the employee (in years)
- `edu`: years of education (in years)
- `startsal`: starting salary (in dollars)
- `jobtime`: time with the company (in months)
- `prevexp`: previous work experience (in months)
- `minority`: yes or no
- `female`: yes or no
- `manager`: yes or no

Read in the data using the code chunk below. 

```{r reademployee}
employee <- read.csv("employee.csv")
```

Preview the data set.

```{r previewemployee}
head(employee)
```

First, let's create a histogram for the `salary` variable. Note: this graph is for the *population* of all employees at this company.

```{r histSalary}
hist(employee$salary,
     xlab = "Salary ($)",
     main = "Histogram of Salaries")
```

The distribution of salaries is unimodal and *heavily skewed to the right*. The center is around 30,000 dollars with the data falling roughly between 10,000 and 140,000 dollars. There are some potential outliers in the upper tail. 

We could also calculate specific numerical summaries. 

```{r summarySalary}
mean(employee$salary)
sd(employee$salary)
```

Think About It: Are these numerical summaries parameters or statistics? 


### Random Samples and Sample Means

R has a few functions that help us draw random samples, but they can be a little bit tricky -- so we have created our own function called `randomSample()`. 

The function takes the following arguments:

- `data`: the name of the data set
- `n`: the desired size of the sample 
- `column`: the column number for the variable of interest

The function will return:

- the variable name
- the values of the random sample
- the mean of the sample

Note: if you get an error that says: could not find function "randomSample", be sure to run the code chunk near the top of the Lab Tutorial (around line 40). 

```{r sampleSalary}
randomSample(data = employee, n = 5, column = 1)
```

**Think-Pair-Share #1:** Consider the following two questions:

- First, run the code chunk above once. Is your sample mean the same as the population mean ($34,418)? Did you expect it to be? Why or why not?
- Then, run the code chunk multiple times, paying close attention to the sample mean value and how much it varies from sample to sample. Do you think it would vary by more, less, or the same if we increased the size of our random sample to 25? Why?


Let's find out what actually happens to the sample mean when we increase the size of our sample to 25. 

**Demo #1**: Use the `randomSample()` function to draw a random sample of size `n = 25` for the `salary` variable (`column = 1`).   

```{r demo1, error = T}
# Replace this text with your code

```

Think About It: Run the code chunk multiple times - paying attention to the value of the sample mean. Does the sample mean seem to vary by more, less, or the same amount (compared to when the sample size was 5)?

Finally, let's see what happens to the sample mean when we increase the size of our sample to 100.  

```{r sampleSalary100}
randomSample(data = employee, n = 100, column = 1)
```

If you run the above code chunk multiple times, you should notice that the sample mean does not vary by much from sample-to-sample!


### The Sampling Distribution of Xbar

A sampling distribution helps us understand, quantify, and visualize the variability of a statistic (like what we are seeing above with the sample mean)! 

A sampling distribution is the distribution of all possible values of a statistic for many, many random samples of the same size (n).

**Note: this is a theoretical distribution.** In reality, we typically only have a *single* sample from the population of interest and make conclusions based on that *one* sample. We know (and have seen) that a statistic will vary from sample to sample. Therefore, it is important to understand how the statistic varies from sample to sample so that we can judge whether our *one* observed statistic is "unusual" or not. 

We can simulate the idea of a sampling distribution using the following steps. Note: for this lab, we will be focusing on the *sample mean* as our statistic of interest.

1. Take a random sample of size n from the population
2. From this random sample, compute the sample mean
3. Save the value of the sample mean computed in Step 2
4. Repeat Steps 1 - 3 many, many times 
5. Create a histogram of the many, many saved sample means

Our first function generated *one* random sample every time we ran the function. What if we wanted to generate *many, many* random samples - say 1000 random samples? We can use the `samplingDistXbar()` function that we created for you! 

The function takes the following arguments:

- `data`: the name of the data set
- `n`: the desired size of the sample 
- `column`: the column number for the variable of interest
- `theoretical`: when set to TRUE, the theoretical distribution is added to the plot (optional)

The function will return a plot of 1000 sample means calculated from 1000 samples of size n. 

We will start by taking samples of size 5. Although our sample size is 5, the function is:

- taking many, many random samples of size 5
- calculating the sample mean for each random sample of size 5
- saving the value for each of the resulting sample means
- creating a plot of these saved sample means

```{r samplingDist5}
samplingDistXbar(data = employee, n = 5, column = 1)
```

Remember the original population distribution for salary (all the way up near Line 100)? Let's compare our sampling distribution (plotted above) to the distribution of the original population:

- The center is: the same (roughly 30,000 - 35,000)
- The spread is: narrower (overall range has decreased)
- The shape is: skewed right, but less severe than the original population


What happens to the sampling distribution when the sample size is increased to 25? 

**Demo #2**: Using the `samplingDistXbar()` function, create the sampling distribution of sample means for random samples of size 25 (`n = 25`) for the `salary` variable (`column = 1`) in the `employee` data set. 

```{r demo2, error = T}
# Replace this text with your code

```

Compared to the previous sampling distribution (when n = 5), you should notice that the center is the same (roughly 30,000 - 35,000) and the spread is even narrower (the overall range has decreased even more).  

Think About It: How would you describe the *shape* of this sampling distribution (when n = 25)? What theorem is being illustrated by this example?


Finally, let's see what happens to the sampling distribution when we increase the size of our sample to 100. 

```{r samplingDist100}
samplingDistXbar(data = employee, n = 100, column = 1)
```

Compared to the previous sampling distributions (when n = 5 and n = 25)

- The center is: the same (roughly 30,000 - 35,000)
- The spread is: even narrower (overall range has decreased even more)
- The shape is: (approximately) normal, even more bell-shaped than Demo #2


**Think-Pair-Share #2:** Consider the following two questions:

- In which of the three scenarios (n = 5, n = 25, n = 100) would the sample mean be the "best" estimator of the true population mean? Why?
- In which of the three scenarios (n = 5, n = 25, n = 100) would observing a sample mean salary of more than $40,000 be the *most* unusual (or unlikely)? Why?



### Conclusions

The above examples helped us illustrate some very important concepts for sample means, all of which can be found on your formula card. 

**Result #1**: The expected value (or center) of the sampling distribution for the sample mean (E(Xbar)) is equal to the mean of the population distribution (mu). In each of the graphs, we saw that the center was 34,418 dollars.

**Result #2**: The standard deviation (or spread) of the sampling distribution for the sample mean (sd(Xbar)) is equal to the population standard deviation divided by the square root of the sample size (sigma / sqrt(n)). We saw that the spread of the sampling distribution decreased as the sample size increased. 

**Result #3**: The shape of the sampling distribution for the sample mean was approximately normal when the sample size was large enough. This last conclusion has a specific name -- the **Central Limit Theorem** -- and is one of the most important concepts in statistics. With a large enough sample size, the distribution of the sample mean will be approximately normal even if the underlying population distribution is not normal. And as we increase the sample size, the normal approximation to the distribution of the sample mean gets better and better.

Note: If the population distribution follows a normal distribution, then the sampling distribution of the sample mean will also follow a normal distribution *regardless of sample size*. 


### Additional Example (Optional)

Let's analyze one more variable - job time (the number of months with the company). First, let's calculate the population mean and population standard deviation. (Remember, these would be considered parameters.)

```{r summaryedu}
mean(employee$jobtime)
sd(employee$jobtime)
```

The population mean (mu) is 81.14 months and the population standard deviation (sigma) is 10.05 months. What about the shape of this distribution? Let's create a histogram to visualize it. 

```{r histedu}
hist(employee$jobtime,
     main = "Histogram of Job Time",
     xlab = "Time with Company (in months)")
```

The distribution is not bell-shaped, but it isn't severely skewed either. 

Run the code chunk below to simulate the sampling distribution of Xbar for various sample sizes. Note: to save space, the following graphs will not show up in your knitted document.

```{r samplingDistEducation}
samplingDistXbar(employee, n = 5, column = 5)
samplingDistXbar(employee, n = 25, column = 5)
samplingDistXbar(employee, n = 100, column = 5)
samplingDistXbar(employee, n = 200, column = 5)
```

These graphs once again highlight the three results from above. They are all centered around 81.14 months, the range of sample means gets narrower as the sample size increases, and the shape of the sampling distribution becomes more and more normal. 

Click on the file titled `lab03assignment.Rmd` in the bottom right window to open up the lab assignment for today. Please do not hesitate to ask questions during lab today - sampling distributions are one of the toughest concepts we cover!

